\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-16 metatheory.tex
Save.

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic an extensional type theoretical framework (with Agda-like syntax), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Our work can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We use as the formulation the intrinsically-typed debruijn syntax in this type-theoretical QIIT instead of a more conventional and untyped representation, because (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a proof because proof checking then is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real theorem prover so that the reader can have an idea about how mechanized formalization can be carried out (5) We can still use named-variables for intuition and example illustration (6) using QIIT as syntax encoding can avoid the reader from the concept of \textit{presupposition}~\citet{gratzer-sterling-birkedal:2019}---these \textit{presupposition} can be understood as \textit{implicit parameters} of the declaration of the judgemental rules in QIIT.

@@

\subsection{Syntax}\YZ{Rewrite syntax part into name-variable style instead of currently debruijn indices, and do that for all the models we introduced}

>>>>

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic an extensional type theoretical framework (with Agda-like syntax), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Our work can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We use as the formulation the intrinsically-typed debruijn syntax in this type-theoretical QIIT instead of a more conventional and untyped representation, because (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a proof because proof checking then is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real theorem prover so that the reader can have an idea about how mechanized formalization can be carried out (5) We can still use named-variables for intuition and example illustration (6) using QIIT as syntax encoding can avoid the reader from the concept of \textit{presupposition}~\cite{gratzer-sterling-birkedal-2019}---these \textit{presupposition} can be understood as \textit{implicit parameters} of the declaration of the judgemental rules in QIIT.

@@

\subsection{Syntax in Declarative Style}
% \YZ{Rewrite syntax part into name-variable style instead of currently debruijn indices, and do that for all the models we introduced}

### 2022-07-16 syntax.tex
Save.

\subsubsection{Warm Up}
\textit{Quotient Inductive Inductive Type}~(QIIT) is an extension of \textit{inductive-inductive type} (IIT)~\cite{nordvall2010inductive}. IIT can be roughly considered as several inductive type mutually defined, in a more expressible way than mutual-inductive type. QIIT is allowing to have \textit{quotient}, in other words, equality results as part of the constructors. For example, in \mintinline{Agda}{data N2: Set where O : N2 | S : N2 → N2 | B : S (S x) ≡ x }, "B" will be a valid (equality) constructor, and "N2" is a set of two elements in our extensional setting.

We highlight some special features of declarative style in QIIT that differ from a
conventional setting: 
(1) This declarative formulation of type theory does not
have an operational semantics but only judgemental equality---$\beta$-reduction is now $\beta$-equality.\footnote{Later we will prove
canonicity which can imply a certain level of computational ability of
our theory} 
(2) We directly use the equality from QIIT to represent judgemental equality in our syntax, relieve from us the burden to explicit mentioning equivalence closure; however, this also mean that, \textit{every piece of the syntax data}, including all kinds of judgements (QIIT type) and their derivations will have equality information upon them  
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
(3) We are dealing with quotient data, so we are actually manipulating
\textit{equivalence classes of terms} instead of one representative (term) from the equivalence class.
(4) Instead of using meta-level substitution, we use explicit
substitution and de Bruijn indices in our intrinsic formulation. This is
also called substitution calculus in the literature and is favoured due
to its closedness to categorical semantic. We will go through these four details one-by-one.

\textbf{Judgements}
We formalize our language as a deep embedding in a type theory with QIIT types. Every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms in MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing} and thus a predicative MLTT, but for readability we
omit the universe levels.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the (pseudo) Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their (pseudo-)Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just as
a well-formed QIIT type is not necessarily inhabited. These well-formedness rules are direct translations of the \textit{well-typedness} of QIIT.

Due to our intrinsic setting, \textbf{we can and will omit a lot of premises of the typing rules}. This can work because we are using
QIIT and everything is intrinsically well-formed(well-typed). For example, we don't
spell out that the rule \ruleref{Type Universe} requires a well-formed
context, even though it does implicitly, \textbf{because it is not
possible to have ill-formed(ill-typed) contexts in our intrinsic setting}. This is
a bit unusual for conventional extrinsic setting where they will put
well-formedness of context everywhere in a lot of term judgements. We
consider this as one advantage of using intrinsic setting---this
``well-formed context'' condition is usually attached to some other premises thus
inferrable and can be considered as ``boilerplate code''. Free of
such boilerplate code, our intrinsic setting provides a more concise
representation.

Quotient is used to represent definitional equality (aka judgemental
equality) between QIIT types concisely.
This way, for all four kinds of judgements we are
using, there is always equality among these data. Doing so, we can omit
some obvious equality judgements. Most of the reasoning can be formulated
using the \textit{algebra} of QIIT (mapping out function from this
quotient data-type).

We quickly review the substitution calculus.
Substitution is very (if not the most) important because function is
first-class citizens, and can be applied (almost) everywhere and can return
(almost) everything, thus we require \textbf{the resulting substitution to
commute} with most (if not all) type/term constructs.

@@

The core language builds on MLTT and adds four new judgement forms.
Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is just a
shorthand for term judgement
$\goodTerm{\Gamma , \cC \sigma}{\_}{A[\pi_1]}$ where $A$ is a type.

@@

Because in dependent type, function now can return type, of which the type is ``type of type'', \textit{the universe} $\cU$. We follow the style of \textit{type universe à la Tarski}~\cite{hofmann1997syntax} and
distinguish types from their \textit{codes or names}, and thus we use
$El\ T$ as the type given \textit{type name} $T : \cU$.
In this case, (1) $\cU$ is the type including the \textit{codes or
names} of types, and (2) $T$ is not allowed to locate at type position
(i.e., at the rightmost position of a term judgement) but $El \ T$ is.
"c" is the ``inverse'' of $El$, obtaining  the \underline{"c"}ode of a
type, with the following two equations $"c" (El~T) \equiv T$ and $El
("c"~T) \equiv T$.
Recall that we are actually using predicative universe, so we actually
have different levels of universes $\cU_j$.
We omit universe levels for conciseness.

@@

$$
\quad 
\Rule[name=Func/DPair Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}
\quad \goodType{\Delta}{A}{} 
\quad \goodType{\Delta, A}{B}{}
}
{
  \goodType{\Gamma}{(\Pi A B)[\gamma] \equiv \Pi A[\gamma] B[\gamma^\uparrow] }{j \cup k}
  \quad 
  \goodType{\Gamma}{(\Sigma A B)[\gamma] \equiv \Sigma A[\gamma] B[\gamma^\uparrow] }{j \cup k}
}
$$
%\YZ{What is the requirement on Delta?}
%\EDJreply{ Delta here is of type Con and thus Delta here should be a well-typed Context. This is another issue I want to talk about, in Agda these kinds of stuff is inferred (so the reader can infer them as well). For example, we know delta is at the place of context, so it is a context, but we don't need to specify that it is a ``well-typed context'' because every context is well-typed in this intrinsic-typing. What's more, I think adding this extra judgement here just cause significant code bloat (like what I complained to you before I know how to use implcit variable in agda. For example, TYPE SUBST, there will be two more judgements (with no more information there) ) (Another reason I prefer (pseudo) agda formulation) }\YZreply{I was thinking about 'Delta |- A'. I suppose it is also implicitly enforced via intrinsically typed syntax. Seems to me it is worth stating, however, or otherwise the rule appears to say that Delta can be any context as long as Gamma |- gamma : Delta.}\EDJreply{You are absolutely right! I think Delta |- A, even though inferrable, mentioning it is good, also some information are B is also ok, please check, resolve if ok, thanks!}
$$
\Rule[name=Type Subst]
{\goodType{\Delta}{T}{j} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{j}}
\quad
\Rule[name=Base Type Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}}
{\goodType{\Gamma}{\cU[\gamma] \equiv \cU }{j + 1} \quad
  \goodType{\Gamma}{\cB[\gamma] \equiv \cB}{0} \quad 
  \goodType{\Gamma}{\bot[\gamma] \equiv \bot}{0}
}
$$
\judgebox{ \goodTerm{\Gamma}{t}{T} }
$$
\Rule[]
{\goodType{\Gamma}{T}{j}}
{\goodTerm{\Gamma}{"c" \ T}{\cU}
}\quad
\Rule[]
{\goodTerm{\Gamma}{T}{\cU}}
{\goodType{\Gamma}{El \ T}{j}}
\quad
\Rule
{}
{\goodTerm{\Gamma}{"tt", "ff"}{\cB}}
\quad 
\Rule[name=Term Subst]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
$$
$$
\Rule[]
{\goodTerm{\Gamma, A}{t}{B}}
{\goodTerm{\Gamma}{\lambda t}{\Pi A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{u}{A} 
\quad \goodTerm{\Gamma}{v}{B[(id, u)]}}
{\goodTerm{\Gamma}{(u,v)}{\Sigma A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{t}{\Sigma A B}}
{\goodTerm{\Gamma}{"pjl" \ t}{A}
\quad \goodTerm{\Gamma}{"pjr" \  t}{B[(id, "pjl" \  t)]}
}
$$
$$
\Rule
{}
{\goodTerm{\Gamma}{(\lambda t)[\gamma] \equiv \lambda (t[\gamma^\uparrow])}{\Pi A B}
\quad \goodTerm{\Gamma}{(u,v)[\gamma] \equiv (u[\gamma],v[\gamma])}{\Sigma A B}
\quad \goodType{\Gamma}{El \ (T[\gamma]) \equiv (El \ T) [\gamma]}{}
}
$$

>>>>

\subsubsection{\textbf{Warm Up}}
\textit{Quotient Inductive Inductive Type}~(QIIT) is an extension of \textit{inductive-inductive type} (IIT)~\cite{nordvall2010inductive}. IIT can be roughly considered as several inductive type mutually defined, in a more expressible way than mutual-inductive type. QIIT is allowing to have \textit{quotient}, in other words, equality results as part of the constructors. For example, in \mintinline{Agda}{data N2: Set where O : N2 | S : N2 → N2 | B : S (S x) ≡ x }, "B" will be a valid (equality) constructor, and "N2" is a set of two elements in our extensional setting.

We highlight some specialities of declarative style in QIIT that differ from a
conventional setting: 
(1) This declarative formulation of type theory does not
have an operational semantics but only judgemental equality---$\beta$-reduction is now $\beta$-equality.\footnote{Later we will prove
canonicity which can imply a certain level of computational ability of
our theory} 
(2) We directly use the equality from QIIT to represent judgemental equality in our syntax, relieve from us the burden to explicit mentioning equivalence closure; however, this also mean that, \textit{every piece of the syntax data}, including all kinds of judgements (QIIT type) and their derivations will have equality information upon them  
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
% (3) We are dealing with quotient data, so we are actually manipulating
% \textit{equivalence classes of terms} instead of one representative (term) from the equivalence class.
(3) Instead of using meta-level substitution, we use explicit
substitution and de Bruijn indices in our intrinsic formulation. This is
also called substitution calculus in the literature and is favoured due
to its closedness to categorical semantic. We will go through these details one-by-one.

\underline{\textit{\textbf{Judgements and Universe.}}}
% Judgements in QIIT
We formalize our language as a deep embedding in a type theory with QIIT types. Every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms from (vanilla) MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing} and thus a predicative MLTT, but for readability we
omit the universe levels.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the (pseudo) Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their (pseudo-)Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just as
a well-formed QIIT type is not necessarily inhabited. 
These well-formedness rules are direct translations of the \textit{well-typedness} of QIIT.

% omitting presupposition
Due to our intrinsic setting, \textbf{we can and will omit a lot of premises of the typing rules}~(the presuppositions). 
For example, we don't explicitly
spell out that the rule \ruleref{Type Universe} (that is stating a universe is a proper type) requires a well-formed
context, even though it does implicitly inferrable, \textbf{because it is not
possible to have ill-formed(ill-typed) contexts in our intrinsic setting}. This is
a bit unusual for conventional extrinsic setting where they will have
well-formedness of context everywhere in most of the term judgements. We
consider this as one advantage of using intrinsic setting---this
``well-formed context'' condition is usually attached to some other premises thus
inferrable and can be considered as ``boilerplate code''. Free of
such boilerplate code, the intrinsic setting provides a more concise
representation.

% How judgemental equality is used
Quotient is used to represent definitional equality (aka judgemental
equality) between QIIT types concisely.
This way, for all four kinds of judgements we are
using, there is always equality among these data but only those equalities between terms and types are emphasized due to their non-triviality.

% Example : how judgemental equality is used for the universe
We review the concept of ``Universe'' and at the same time showing how equality is representing semantic. In dependent type setting, function now can return type, of which the type is ``type of type'', i.e., \textit{the universe} $\cU$. We follow the style of \textit{type universe à la Tarski}~\cite{hofmann1997syntax} and
distinguish types from their \textit{codes or names} as showin in \ruleref{Codes/Get Type}: given a \textit{code} (actually a term) $\goodTerm{\Gamma}{T}{\cU}$ we can retrieve the corresponding type
$\goodType{\Gamma}{El~T}{}$, and vice-versa (given a

In other words, (1) a universe $\cU$ is a type including the \textit{codes or names} of types.
(2) a code (term) $\goodTerm{\Gamma}{T}{\cU}$ is not allowed to locate at type position
(i.e., at the rightmost position of a term judgement) but $El \ T$ is (i.e. we distinguish the code and the type). (3)
"c" is the ``inverse'' of $El$, obtaining  the \underline{"c"}ode of a
type, with the following \textit{two (omitted) equations representing their semantic} $"c" (El~T) \equiv T$ and $El
("c"~T) \equiv T$---notice how the two sides of each equality is term of the same type thus well-typed. More concretely,

Note that we are actually using predicative universe, so we actually
have different levels of universes $\cU_j$. However, we omit universe levels in our latex presentation for conciseness and accessibility.
% Most of the reasoning can be formulated using the \textit{algebra} of QIIT (mapping out function from this quotient data-type).

\underline{\textit{\textbf{Substitution Calculus.}}}
We quickly review the substitution calculus.
Substitution is very (if not the most) important because (dependent) functions are
first-class citizens, and lambdas can wrap (almost) every parts and can return
(almost) everything, thus we require \textbf{the resulting substitution to
commute} with most (if not all) type/term constructs.

@@

\subsubsection{\textbf{Core Syntax of the Extended MLTT}}
The core language builds on MLTT and adds four new judgement forms.
Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is just a
shorthand for term judgement
$\goodTerm{\Gamma , \cC \sigma}{\_}{A[\pi_1]}$ where $A$ is a type.

@@

@@

$$
\quad 
\Rule[name=Func/DPair Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}
\quad \goodType{\Delta}{A}{} 
\quad \goodType{\Delta, A}{B}{}
}
{
  \goodType{\Gamma}{(\Pi A B)[\gamma] \equiv \Pi A[\gamma] B[\gamma^\uparrow] }{j \cup k}
  \quad 
  \goodType{\Gamma}{(\Sigma A B)[\gamma] \equiv \Sigma A[\gamma] B[\gamma^\uparrow] }{j \cup k}
}
$$
%\YZ{What is the requirement on Delta?}
%\EDJreply{ Delta here is of type Con and thus Delta here should be a well-typed Context. This is another issue I want to talk about, in Agda these kinds of stuff is inferred (so the reader can infer them as well). For example, we know delta is at the place of context, so it is a context, but we don't need to specify that it is a ``well-typed context'' because every context is well-typed in this intrinsic-typing. What's more, I think adding this extra judgement here just cause significant code bloat (like what I complained to you before I know how to use implcit variable in agda. For example, TYPE SUBST, there will be two more judgements (with no more information there) ) (Another reason I prefer (pseudo) agda formulation) }\YZreply{I was thinking about 'Delta |- A'. I suppose it is also implicitly enforced via intrinsically typed syntax. Seems to me it is worth stating, however, or otherwise the rule appears to say that Delta can be any context as long as Gamma |- gamma : Delta.}\EDJreply{You are absolutely right! I think Delta |- A, even though inferrable, mentioning it is good, also some information are B is also ok, please check, resolve if ok, thanks!}
$$
\Rule[name=Type Subst]
{\goodType{\Delta}{T}{j} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{j}}
\quad
\Rule[name=Base Type Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}}
{\goodType{\Gamma}{\cU[\gamma] \equiv \cU }{j + 1} \quad
  \goodType{\Gamma}{\cB[\gamma] \equiv \cB}{0} \quad 
  \goodType{\Gamma}{\bot[\gamma] \equiv \bot}{0}
}
$$
\judgebox{ \goodTerm{\Gamma}{t}{T} }
$$
\Rule[name=Code/Get Type]
{\goodType{\Gamma}{T}{j}}
{\goodTerm{\Gamma}{"c" \ T}{\cU}
}\quad
\Rule[]
{\goodTerm{\Gamma}{T}{\cU}}
{\goodType{\Gamma}{El \ T}{j}}
\quad
\Rule
{}
{\goodTerm{\Gamma}{"tt", "ff"}{\cB}}
\quad 
\Rule[name=Term Subst]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
$$
$$
\Rule[]
{\goodTerm{\Gamma, A}{t}{B}}
{\goodTerm{\Gamma}{\lambda t}{\Pi A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{u}{A} 
\quad \goodTerm{\Gamma}{v}{B[(id, u)]}}
{\goodTerm{\Gamma}{(u,v)}{\Sigma A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{t}{\Sigma A B}}
{\goodTerm{\Gamma}{"pjl" \ t}{A}
\quad \goodTerm{\Gamma}{"pjr" \  t}{B[(id, "pjl" \  t)]}
}
$$
$$
\Rule
{}
{\goodTerm{\Gamma}{(\lambda t)[\gamma] \equiv \lambda (t[\gamma^\uparrow])}{\Pi A B}
\quad \goodTerm{\Gamma}{(u,v)[\gamma] \equiv (u[\gamma],v[\gamma])}{\Sigma A B}
\quad \goodType{\Gamma}{El \ (T[\gamma]) \equiv (El \ T) [\gamma]}{}
}
$$
